Supplemental Material for: The evolution of gene expression in the term placenta of viviparous mammals

### 2016-11-30 manuscript/supplemental_material.Rnw
use the housekeeping shape code from the manuscript

\begin{figure}
<<housekeeping.placenta.distribution,fig.width=8,fig.height=8,out.height="0.8\\textheight",echo=FALSE,message=FALSE>>=
## housekeeping set as apposed to non-housekeeping set
cptg.h <- core.placenta.transcriptome.genes.shape[(human_name %in% housekeeping.genes.superset[,gene_short_name])]
num.species <- star.logs.wide[,length(unique(species))]

@@

cpt.h.distribution.shape <- 
    lapply(c("all_expression",
             "all_but_one_expression",
             "all_but_two_expression",
             "all_but_three_expression",
             "all_but_four_expression"),
           function(x){
               calculate.distribution.shape(cptg.h[,x,with=FALSE][[1]])[,`Minimum Expression in`:=x]
           })
fix.expression.names <- function(x){
    x <- gsub("^all_expression$",
              paste0(num.species," of ",num.species," species"),
              x
              )
    nc.levels <- paste0(num.species," of ",num.species," species")
    ## number coutner
    nc <- c("one"=1,"two"=2,"three"=3,"four"=4)
    for (a in names(nc)) {
        x <- gsub(paste0("^all_but_",a,"_expression$"),
              paste0(num.species-nc[a]," of ",num.species," species"),
              x
              )
        nc.levels <-
            c(nc.levels,
              paste0(num.species-nc[a]," of ",num.species," species"))
    }
    return(factor(x,levels=nc.levels))
}

@@

cpt.h.distribution.shape <- 
    cpt.h.distribution.shape[,`Minimum Expression in`:=
                                fix.expression.names(`Minimum Expression in`)]

@@

print(ggplot(cpt.h.distribution.shape[`Expression (FPKM)` > 0.1,],
             aes(x=`Expression (FPKM)`,color=`Minimum Expression in`,y=num_genes))
      + geom_line(lwd=1)
      + xlab("Minimum Expression cutoff (FPKM)")
      + ylab("Genes with N-th Minimum Expression")
      + scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
                      labels = trans_format("log10", math_format(10^.x)))
      + scale_x_log10(breaks = trans_breaks("log10", function(x) 10^x),
                      labels = trans_format("log10", math_format(10^.x)))
      + geom_vline(xintercept=10)
      + geom_segment(mapping=aes(y=V1,
                         yend=V1,
                         xend=10,
                         x=250,
                         color=`Minimum Expression in`),
                     data=cpt.h.distribution.shape[`Expression (FPKM)` > 10,
                                              max(num_genes),`Minimum Expression in`],
                     arrow=arrow(length=unit(0.5,"cm"))
                   )
      + geom_text(mapping=aes(x=250,
                      y=V1,
                      label=paste0(V1," genes"),override=TRUE),
                  data=cpt.h.distribution.shape[`Expression (FPKM)` > 10,
                      max(num_genes),`Minimum Expression in`],
                  show.legend=FALSE,
                  hjust=0
                  )
      + guides(color=guide_legend(title="Minimum\nExpression in"))
      )
@ 
\caption{Distribution shape of housekeeping genes in placentas out of
  \Sexpr{pts[,sum(expressed.housekeeping)]} \human housekeeping 
  genes which are expressed in any placenta. The vertical axis
  shows the number of 1:1 orthologous \human housekeeping genes
  expressed in 
  placenta transcriptomes if %
  \Sexpr{cpt.h.distribution.shape[,levels(`Minimum Expression in`)[1]]}
  %
  (or %
  \Sexpr{cpt.h.distribution.shape[,levels(`Minimum Expression in`)[2]]},
  % 
  etc.) have a minimum expression in FPKM equal to the horizontal axis.
  For example, if the cutoff is set to 10 (vertical black line), 
  % 
  \Sexpr{cpt.h.distribution.shape[`Expression (FPKM)` > 10 
    & `Minimum Expression in` == '14 of 14 species',max(num_genes)]}
  %
  genes would be included if the FPKM of a set of 1:1 orthologs of all 14 species must be at
  least 10, %
  \Sexpr{cpt.h.distribution.shape[`Expression (FPKM)` > 10 
    & `Minimum Expression in` == "13 of 14 species",max(num_genes)]}
  %
  1:1 orthologs would be included if 13 of 14 species have an
   with an FPKM of at least 10, and 
  \Sexpr{cpt.h.distribution.shape[`Expression (FPKM)` > 10 
    & `Minimum Expression in` == '10 of 14 species',max(num_genes)]}
  genes would be included if 10 of 14 species
  must have an FPKM of at least 10.}
\label{fig:core_housekeeping_transcriptome_shape}
\end{figure}

>>>>

\begin{figure}
<<housekeeping.placenta.distribution,fig.width=5,fig.height=5,out.height="0.8\\textheight",echo=FALSE,message=FALSE>>=
## housekeeping set as apposed to non-housekeeping set
cptg.h <- core.placenta.transcriptome.genes.shape[(human_name %in% housekeeping.genes.superset[,gene_short_name])]
num.species <- star.logs.wide[,length(unique(species))]

@@

cpt.h.distribution.shape <- 
    lapply(c("all_expression",
             "all_but_one_expression",
             "all_but_two_expression",
             "all_but_three_expression",
             "all_but_four_expression",
             "all_but_metatherian_expression"
             ),
           function(x){
               calculate.distribution.shape(cptg.h[,x,with=FALSE][[1]])[,`Minimum Expression in`:=x]
           })
fix.expression.names <- function(x){
    x <- gsub("^all_but_metatherian_expression$",
              paste0("All ",num.species-1," Eutherians"),
              x
              )
    x <- gsub("^all_expression$",
              paste0(num.species," of ",num.species," Species"),
              x
              )
    nc.levels <- c(paste0("All ",num.species-1," Eutherians"),
                   paste0(num.species," of ",num.species," Species")
                   )
    ## number coutner
    nc <- c("one"=1,"two"=2,"three"=3,"four"=4)
    for (a in names(nc)) {
        x <- gsub(paste0("^all_but_",a,"_expression$"),
              paste0(num.species-nc[a]," of ",num.species," Species"),
              x
              )
        nc.levels <-
            c(nc.levels,
              paste0(num.species-nc[a]," of ",num.species," Species"))
    }
    return(factor(x,levels=nc.levels))
}

@@

cpt.h.distribution.shape <- 
    cpt.h.distribution.shape[,`Min. Exp. in`:=
                                fix.expression.names(`Minimum Expression in`)]

@@

print(ggplot(cpt.h.distribution.shape[`Expression (FPKM)` > 0.1,],
             aes(x=`Expression (FPKM)`,color=`Min. Exp. in`,y=num_genes))
      + geom_line(lwd=1)
      + xlab("Minimum Expression cutoff (FPKM)")
      + ylab("Number of Genes Meeting FPKM Cutoff")
      + scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
                      labels = trans_format("log10", math_format(10^.x)))
      + scale_x_log10(breaks = trans_breaks("log10", function(x) 10^x),
                      labels = trans_format("log10", math_format(10^.x)))
      + geom_vline(xintercept=10,linetype="dotted")
      + geom_segment(mapping=aes(y=V1-ifelse(V1<145,40,0),
                         yend=V1,
                         xend=10,
                         x=sapply(V1,
                                  function(x){
                                      cpt.h.distribution.shape[num_genes > x,
                                                             max(`Expression (FPKM)`)*2.5]
                                  }),
                         color=`Min. Exp. in`),
                     data=cpt.h.distribution.shape[`Expression (FPKM)` > 10,
                                              max(num_genes),`Min. Exp. in`],
                     arrow=arrow(length=unit(0.5,"cm")),
                     show.legend=FALSE
                     )
      + geom_text(mapping=aes(x=sapply(V1,
                                       function(x){
                                           cpt.h.distribution.shape[num_genes > x,
                                                                  max(`Expression (FPKM)`)*2.6]
                                       }),
                      y=V1-ifelse(V1<145,40,0),
                      label=paste0(V1," genes"),override=TRUE),
                  data=cpt.h.distribution.shape[`Expression (FPKM)` > 10,
                      max(num_genes),`Min. Exp. in`],
                  show.legend=FALSE,
                  hjust=0,
                  size=8*(5/14)
                  )
      + geom_segment(mapping=aes(yend=V1,
                         y=100,
                         xend=1,
                         x=1,
                         color=`Min. Exp. in`),
                     data=cpt.h.distribution.shape[`Expression (FPKM)` > 1 &
                                                 `Min. Exp. in` == 'All 13 Eutherians',
                                                 max(num_genes),`Min. Exp. in`],
                     arrow=arrow(length=unit(0.5,"cm")),
                     show.legend=FALSE
                     )
      + geom_text(mapping=aes(x=1,
                      y=100,
                      label=paste0(V1," genes"),
                      override=TRUE),
                  angle=90,
                  data=cpt.h.distribution.shape[`Expression (FPKM)` > 1 &
                                              `Min. Exp. in` == 'All 13 Eutherians',
                                              max(num_genes),`Min. Exp. in`],
                  show.legend=FALSE,
                  hjust=1,
                  size=8*(5/14)
                  )
      + guides(color=guide_legend(title=paste0("Minimum\nExpression in"),
                                  keywidth=unit(3,"mm"),
                                  keyheight=unit(3,"mm"),
                                  title.hjust=0.5,
                                  label.position="right",
                                  label.theme=element_text(size=7,angle=0)
                                  ))
      + theme_bw()
      + theme(text=element_text(size=9),
              legend.position=c(0.99,0.99),
              legend.margin=unit(0,"mm"),
              legend.justification=c(1,1),
              plot.margin=unit(c(0,0,0,0),"mm")
              )
      )
\caption{Distribution shape of housekeeping genes in placentas out of
  \Sexpr{pts[,sum(expressed.housekeeping)]} \human housekeeping 
  genes which are expressed in any placenta. The vertical axis
  shows the number of 1:1 orthologous \human housekeeping genes
  expressed in 
  placenta transcriptomes if %
  \Sexpr{cpt.h.distribution.shape[,levels(`Min. Exp. in`)[1]]}
  %
  (or %
  \Sexpr{cpt.h.distribution.shape[,levels(`Min. Exp. in`)[2]]},
  % 
  etc.) have a minimum expression in FPKM equal to the horizontal axis.
  For example, if the cutoff is set to 10 (vertical black line), 
  % 
  \Sexpr{cpt.h.distribution.shape[`Expression (FPKM)` > 10 
    & `Min. Exp. in` == '14 of 14 Species',max(num_genes)]}
  %
  genes would be included if the expression of a set of 1:1 orthologs of all 14 species must be at
  least 10 FPKM, %
  \Sexpr{cpt.h.distribution.shape[`Expression (FPKM)` > 10 
    & `Min. Exp. in` == "13 of 14 Species",max(num_genes)]}
  %
  1:1 orthologs would be included if 13 of 14 species have an
   expression of at least 10 FPKM, and 
  \Sexpr{cpt.h.distribution.shape[`Expression (FPKM)` > 10 
    & `Min. Exp. in` == '10 of 14 Species',max(num_genes)]}
  genes would be included if 10 of 14 species
  must have an expression of at least 10 FPKM.}
\label{fig:core_housekeeping_transcriptome_shape}
\end{figure}
